---
title: 'Point Estimation Practice'
format: html
editor: source
---

# Practice Problems

## Method of Moments I

Use the method of moments to construct an estimator for the parameter $N$ when your data $X_1, X_2, \ldots, X_n$ each are drawn independently from $Binom(N, p)$. In this case assume that $p$ is known to you. Note the sample size you're working with is not the same as the $N$ parameter from the binomial distribution.

::: {.callout-note collapse="true" title="Solution"}
Since $EX=Np$, if $p$ is known, and since $\bar{X}\approx EX$ we can create an estimator for $N$ by substituting $\bar X$ for $EX$ and replace the equals with approximation.

$$\bar{X} \approx N p$$ so
$$N \approx \frac{\bar X}{p}$$
:::


## Method of Moments II

Continuing from the previous problem, suppose that your data $X_1, X_2, \ldots, X_n$ each are drawn independently from $Binom(N, p)$ but both $N$ and $p$ are unknown. Use the method of moments to construct two simultaneous estimators $\hat{N}$ and $\hat{p}$. You can only use sample statistics in your estimator functions, you cannot use $N$ or $p$ because both of these parameters are unknown.

::: {.callout-note collapse="true" title="Solution"}
This will require us to use both the first and second moments. In place of the second moment $EX^2$ we can use the variance, with the approximation that $VarX \approx S^2$.

So we can make two approximations relating the sample mean and sample variance:

$$\bar X \approx EX = Np$$
$$S^2 \approx VarX = Np(1-p)$$
We can take a ratio to remove the $N$:
$$\frac{Var X}{\bar X} \approx \frac{Np(1-p)}{Np}=1-p$$
Solve for $p$ and we get $$\hat p = 1-\frac{Var X}{\bar X}$$

With this, we can find the estimator for $N$

$$\bar{X} \approx Np \Rightarrow N \approx \frac{\bar{X}}{p}$$
we can substitute in $\hat p$
$$\hat N = \frac{\bar{X}}{\hat p}=\frac{\bar X}{1-VarX/\bar{X}}=\frac{\bar{X}^2}{\bar X - Var X}$$
:::


## Method of Moments III

Suppose $X_1, \ldots, X_n \sim Unif(0, M)$. Use the method of moments to construct an estimator $\hat{M}$. 

::: {.callout-note collapse="true" title="Solution"}
Since $EX=\frac{M+0}{2}$, We can say $\bar{X}\approx \frac{M}{2}$
$$\hat M = 2\bar{X}$$
:::


## Method of Moments IV

The gamma distribution takes two parameters, the shape parameter $\alpha>0$ and the scale parameter $\theta>0$. For $X\sim Gamma(\alpha, \theta)$, we have two properties of the distribution: $EX=\alpha\theta$ and $VarX=\alpha\theta^2$. Suppose we have a random sample $X_1, \ldots, X_n$ from such a gamma distribution.

a. If you know $\theta$, use the method of moments to find an estimator $\hat\alpha$.

::: {.callout-note collapse="true" title="Solution"}

Since $\bar X \approx EX = \alpha\theta$, if $\theta$ is known, then we can simply use $$\hat\alpha = \frac{\bar{X}}{\theta}$$
:::


b. If you do not know $\theta$, use the method of moments to find an estimator $\hat\alpha$. Note this estimator will need to use both sample mean and sample variance in some way.

::: {.callout-note collapse="true" title="Solution"}
Now we must invoke the approximation $$S^2 \approx Var X = \alpha\theta^2$$
If we square the expected value we get $$E^2X = \alpha^2\theta^2$$
The ratio will be $\alpha$:
$$\frac{E^2X}{Var X}=\frac{\alpha^2\theta^2}{\alpha\theta^2}=\alpha$$
Thus $$\hat\alpha = \frac{\bar{X}^2}{S^2}$$
:::


## Method of Moments V

Suppose we observe a sequence of independent trials where each trial is a success or a failure. Let $X$ denote the number of failures before the first success. This experiment is repeated $n$ times. Use the method of moments to construct an estimator for $p$, the probability of success.

::: {.callout-note collapse="true" title="Solution"}
This describes a geometric random variable, as defined in STAT 340, and how it is defined in R. If $X \sim Geometric(p)$, $EX=\frac{1-p}{p}$

Thus we could make an approximation

$$\bar X \approx \frac{1-p}{p}$$

$$
\begin{align}
\bar X &\approx \frac{1-p}{p}\\
p\bar{X} &\approx 1-p \\
p\bar{X}+p &\approx 1\\
p(\bar{X}+1) &\approx 1\\
p &\approx \frac{1}{\bar{X}+1}
\end{align}
$$
So this is our estimator
$$\hat p  = \frac{1}{\bar{X}+1}$$
:::


## Bias Sample SD

Suppose we have a normally distributed population, and an independent sample of size $15$ from this population. So our data $X_1,\ldots,X_{15}\sim N(10, \sigma^2)$. For parts (a) and (b) you can set $\sigma$ to be whatever value you want.

a. Use Monte Carlo to demonstrate that $S^2$ is an unbiased estimator of $\sigma^2$. 

::: {.callout-note collapse="true" title="Solution"}

Let's generate a bunch of samples and show all of the estimates for $\sigma^2$ on a plot.

```{r}
# Let sigma = 5
sampleVars <- replicate(10000, var(rnorm(15, 10, 5)))
hist(sampleVars)
abline(v=25, col="red")
abline(v=mean(sampleVars), lty=2)

```

:::

b. Use Monte Carlo to demonstrate that $S$, the sample standard deviation, is a *biased* estimator of $\sigma$, the population variance. 

::: {.callout-note collapse="true" title="Solution"}

Let's generate a bunch of samples and show all of the sample standard deviations on a plot.

```{r}
# Let sigma = 5
sampleSDs <- replicate(10000, sd(rnorm(15, 10, 5)))
hist(sampleSDs)
abline(v=5, col="red")
abline(v=mean(sampleSDs), lty=2)

```

:::

## A Poisson Estimator
Your random sample $X_1, \ldots, X_20 \sim Poisson(\lambda)$. For parts (a) and (b) let's assume the true value of $\lambda$ is 5.

a. Use Monte Carlo to demonstrate that $\bar{X}$ and $S^2$ are both unbiased estimators of $\lambda$.

::: {.callout-note collapse="true" title="Solution"}

Let's generate a bunch of samples and show all of the estimates for $\sigma^2$ on a plot.

```{r}
# Let sigma = 5
sampleVars <- replicate(10000, var(rpois(20, 5)))
sampleMeans <- replicate(10000, mean(rpois(20, 5)))
par(mfrow=c(1,2))
hist(sampleVars, breaks=40)
abline(v=5, col="red")
abline(v=mean(sampleVars), lty=2)
hist(sampleMeans, breaks=40)
abline(v=5, col="red")
abline(v=mean(sampleMeans), lty=2)
```

:::


b. Which estimator $\bar{X}$ or $S^2$ has a lower variance (i.e. is a more precise) estimator for $\lambda$? Use Monte Carlo support your answer.

::: {.callout-note collapse="true" title="Solution"}
We already have samples so we can just calculat ehte sample variances
```{r}
var(sampleVars)
var(sampleMeans)
```

$\bar{X}$ clearly has a much lower variance.

:::


## The Tricky Sample Maximum

Suppose $X_1, \ldots, X_n \sim Unif(0, M)$ (continuous). Consider two estimators for $M$: 

$$\hat{M}_1 = \max(X_1, \ldots, X_n)$$
$$\hat{M}_2 = \frac{n+1}{n}\max(X_1, \ldots, X_n)$$
Suppose that You have a sample size $n=25$ and a true $M=100$.

a. Use Monte Carlo  to estimate the bias of each of these estimators.

::: {.callout-note collapse="true" title="Solution"}

```{r}
M1 <- 0; M2<-0
for(i in 1:100000){
  x <- runif(25, 0, 100)
  M1[i] <- max(x)
  M2[i] <- 26/25*max(x)
}
mean(M1)-100
mean(M2)-100

```
The first estimator seems to be negatively biased, but the second seems to be approximately unbiased.
:::

b. Use Monte Carlo to estimate the variance of both of these estimators. 

::: {.callout-note collapse="true" title="Solution"}

```{r}
var(M1)
var(M2)
```
M1 (which is biased) has a smaller variance M2 is unbiased and has a larger variance. 
:::

c. The Mean Squared Error is $E\left[(\hat\theta-\theta)^2\right]$. This can be approximated by Monte Carlo by estimating the parameter repeatedly, squaring the error and averaging. Use Monte Carlo to estimate the MSE of both $\hat{M}_1$ and $\hat{M}_2$.

::: {.callout-note collapse="true" title="Solution"}

```{r}
mean((M1-100)^2)
mean((M2-100)^2)

```
The second estimator, $\hat M_2$ has a lower mean squared error.
:::

d. Consider the estimator $$\hat{M}_k=\frac{n+k}{n}\max(X_1,\ldots,X_n)$$ for values of $k \in [-1,3]$. For which value of $k$ is the MSE minimized? Can you plot the squared bias, variance and MSE on one plot?

::: {.callout-note collapse="true" title="Solution"}
We can just get 10000 sample maximums and construct these estimators easily. Since I've already got my `M1` object, I can use that.

```{r}
n <- 25
bias2 <- 0
variance <- 0
mse <- 0
k <- seq(-1,3, by=.1)

for(i in 1:length(k)){
  bias2[i]=(mean((n+k[i])/n*M1)-100)^2
  variance[i]=var((n+k[i])/n*M1)
  mse[i] = mean(((n+k[i])/n*M1-100)^2)
}

plot(x=k, y=variance, col="red",type="l", ylim=c(0,max(c(bias2,variance,mse))), ylab="")
lines(x=k, y=bias2, col="blue")
lines(x=k, y=mse)
legend(x="top", legend=c("variance","squared bias","mse"), col=c("red","blue","black"), lty=1)

k[which.min(mse)]

```
It seems the $k$ value that minimizes the mean squared error is $k=1$.
:::


## Tricky Sample Maximum II

Another couple estimators for $M$ from the previous problem are $\hat{M}_3=2\bar{X}$, twice the sample mean and $\hat{M}_4=2\tilde{X}$, twice the sample median. 

Suppose that You have a sample size $n=25$ and a true $M=100$.

a. Use Monte Carlo to estimate the bias and variance of each estimator.

::: {.callout-note collapse="true" title="Solution"}

```{r}
M3 <- 0; M4<-0
for(i in 1:100000){
  x <- runif(25, 0, 100)
  M3[i] <- 2*mean(x)
  M4[i] <- 2*median(x)
}
#biases
mean(M3)-100
mean(M4)-100
#variances
var(M3)
var(M4)
```
M3 has a slight negative bias, M4 seems to be more negatively biased. The variance of M4 is also higher than the variance of M3.

:::

b. Use Monte Carlo to approximate the MSE of each estimator.

::: {.callout-note collapse="true" title="Solution"}
```{r}
mean((M3-100)^2)
mean((M4-100)^2)
```
Both of these estimators have a far worse MSE than either of the first two.
:::

c. Between $\hat{M}_1$ through $\hat{M}_4$, which estimator would you prefer? What are some pros and cons of each of them?

::: {.callout-note collapse="true" title="Solution"}
Between all four estimators, M2 definitely has the smallest MSE and it is an unbiased estimator. It's definitely the preferred estimator. 
:::

## Exponential Data

Consider the data below given by Wang (2000) on failure times of an electrical component. Assuming an exponential distribution $Exp(\lambda)$, find the method of moments estimate of $\lambda$

```{r}
t <- c(5, 11, 21, 31, 46, 75, 98, 122, 145, 165, 196, 224, 245, 293, 321, 330, 350, 420)
```

::: {.callout-note collapse="true" title="Solution"}
For an exponential random variable $X$, $EX=\frac 1\lambda$. The method of moments suggests that $\bar {X} \approx \frac1\lambda$, so $\hat\lambda = \frac1{\bar{X}}$

```{r}
1/mean(t)
```
We'd estimate $\hat\lambda=.0058102$.
:::




## Another Uniform Estimation

Suppose $X_1, \ldots X_n \sim Unif(-\theta, theta)$ are an independent random sample. 

a. Use the method of moments to construct an estimator for $\theta$ (hint: the first moment will be useless!)

::: {.callout-note collapse="true" title="Solution"}
This is a fun one. Because $EX={\frac{a+b}{2}}$, in this case that would be $EX=\frac{-\theta + \theta}{2}=0$. The second moment (well, technically we'll use the variance) is actually useful. $VarX=\frac{(b-a)^2}{12}=\frac{(\theta- -\theta)^2}{12}=\frac{\theta^2}{12}$

So the method of moments approximates $VarX$ with $S^2$: $$S^2\approx \frac{\theta^2}{12} \Rightarrow \theta \approx \sqrt{12}S$$

So our method of moments estimator would be $$\hat\theta = \sqrt{12}S$$

:::

b. Construct another estimator based on the sample max and sample min. Is it biased?

::: {.callout-note collapse="true" title="Solution"}

An intuitive estimator based on these two  sample stats ($M_{(1)}$ the sample minimum, and $M_{(n)}$ the sample maximum) would be $\hat\theta_2=\frac{M_{(n)}-M_{(1)}}{2}$, but this is probably going to underestimate $\theta$.

```{r}
#Let theta be 1, and n=20
theta=1
ranges <- replicate(10000, diff(range(runif(20, -theta,theta))))
mean(ranges/2)
```
Obviously we have underestimated $\theta$. It turns out that an unbiased estimator for the range of a uniform distribution is $\frac{n+1}{n-1}(\max(X)-\min(X))$. Since The range is $2\theta$, if we divide this estimate by 2 we should get an unbiased estimate of $\theta$.

```{r}
mean(((20+1)/(19)*ranges)/2)

```

:::

## Goldfish

In 1997 Pepperidge Farm started putting faces on some of the goldfish crackers they produce. Roughly 40% of goldfish crackers have faces.

Suppose you want to estimate this proportion, and you buy a 20 pack box as a representative sample of Goldfish. The box contains 20 1-oz packs of goldfish, each containing roughly 50 goldfish. Suppose we can model the number of goldfish in a pack using a normal distribution (rounded to the nearest integer), $Y_i \sim N(50, 2^2)$.

::: {.callout-note collapse="true" title="Solution"}

To estimate the proportion of smiley faces, you figure you could do three things:

1. Open all 20 bags in one pile and calculate the sample proportion
2. Calculate the proportion in each of the 20 bags and average those proportions
3. Calculate the proportions in each bag and calculate the **weighted** average (weighting by the number of goldfish in each bag)

Using a normal model for the number of fish in a bag, and a binomial for the number of smiley fish, determine the bias and variance of these three estimators.

```{r}
NMC <- 10000
estimates <- data.frame(method1=numeric(NMC),
                        method2=numeric(NMC),
                        method3=numeric(NMC))
for(i in 1:NMC){
  Y <- round(rnorm(20, 50, 2))
  X <- rbinom(20, Y, .40)
  estimates[i,] <- c(sum(X)/sum(Y),
                     mean(X/Y),
                     sum(X/Y * Y/sum(Y))) 
}

colMeans(estimates)
apply(estimates, 2, FUN=var)
```
When working out the math, it becomes clear that method 1 and 3 are actually identical to each other. 
Running the MC a few times it seems that all 3 methods give an unbiased estimate. Method 2 however has a slightly larger variance. This seems consistent after re-running the MC multiple times. Thus method 1 is probably the best method to estimate $p$.

:::

# Beyond STAT 340

These problems are excellent practice but they are beyond the material we cover in STAT 340.


(none yet)

