---
title: 'Point Estimation Practice'
format: html
editor: source
---

# Practice Problems

## Method of Moments I

Use the method of moments to construct an estimator for the parameter $N$ when your data $X_1, X_2, \ldots, X_n$ each are drawn independently from $Binom(N, p)$. In this case assume that $p$ is known to you. Note the sample size you're working with is not the same as the $N$ parameter from the binomial distribution.

## Method of Moments II

Continuing from the previous problem, suppose that your data $X_1, X_2, \ldots, X_n$ each are drawn independently from $Binom(N, p)$ but both $N$ and $p$ are unknown. Use the method of moments to construct two simultaneous estimators $\hat{N}$ and $\hat{p}$. You can only use sample statistics in your estimator functions, you cannot use $N$ or $p$ because both of these parameters are unknown.

## Method of Moments III

Suppose $X_1, \ldots, X_n \sim Unif(0, M)$. Use the method of moments to construct an estimator $\hat{M}$. 

## Method of Moments IV

The gamma distribution takes two parameters, the shape parameter $\alpha>0$ and the scale parameter $\theta>0$. For $X\sim Gamma(\alpha, \theta)$, we have two properties of the distribution: $EX=\alpha\theta$ and $VarX=\alpha\theta^2$. Suppose we have a random sample $X_1, \ldots, X_n$ from such a gamma distribution.

a. If you know $\theta$, use the method of moments to find an estimator $\hat\alpha$.

b. If you do not know $\theta$, use the method of moments to find an estimator $\hat\alpha$. Note this estimator will need to use both sample mean and sample variance in some way.

## Method of Moments V

Suppose we observe a sequence of independent trials where each trial is a success or a failure. Let $X$ denote the number of failures before the first success. This experiment is repeated $n$ times. Use the method of moments to construct an estimator for $p$, the probability of success.

## Bias Sample SD

Suppose we have a normally distributed population, and an independent sample of size $15$ from this population. So our data $X_1,\ldots,X_{15}\sim N(10, \sigma^2)$. For parts (a) and (b) you can set $\sigma$ to be whatever value you want.

a. Use Monte Carlo to demonstrate that $S^2$ is an unbiased estimator of $\sigma^2$. 

b. Use Monte Carlo to demonstrate that $S$, the sample standard deviation, is a *biased* estimator of $\sigma$, the population variance. 

## A Poisson Estimator
Your random sample $X_1, \ldots, X_20 \sim Poisson(\lambda)$. For parts (a) and (b) let's assume the true value of $\lambda$ is 5.

a. Use Monte Carlo to demonstrate that $\bar{X}$ and $S^2$ are both unbiased estimators of $\lambda$.

b. Which estimator $\bar{X}$ or $S^2$ has a lower variance (i.e. is a more precise) estimator for $\lambda$? Use Monte Carlo support your answer.

## The Tricky Sample Maximum

Suppose $X_1, \ldots, X_n \sim Unif(0, M)$ (continuous). Consider two estimators for $M$: 

$$\hat{M}_1 = \max(X_1, \ldots, X_n)$$
$$\hat{M}_2 = \frac{n+1}{n}\max(X_1, \ldots, X_n)$$
Suppose that You have a sample size $n=25$ and a true $M=100$.

a. Use Monte Carlo  to estimate the bias of each of these estimators.

b. Use Monte Carlo to estimate the variance of both of these estimators. 

c. The Mean Squared Error is $E\left[(\hat\theta-\theta)^2\right]$. This can be approximated by Monte Carlo by estimating the parameter repeatedly, squaring the error and averaging. Use Monte Carlo to estimate the MSE of both $\hat{M}_1$ and $\hat{M}_2$.

## Tricky Sample Maximum II

Another couple estimators for $M$ from the previous problem is $\hat{M}_3=2\bar{X}$, twice the sample mean and $\hat{M}_4=2\tilde{X}$, twice the sample median. 

Suppose that You have a sample size $n=25$ and a true $M=100$.

a. Use Monte Carlo to estimate the bias and variance of each estimator.

b. Use Monte Carlo to approximate the MSE of each estimator.

c. Between $\hat{M}_1$ through $\hat{M}_4$, which estimator would you prefer? What are some pros and cons of each of them?

## Exponential Data

Consider the data below given by Wang (2000) on failure times of an electrical component. Assuming an exponential distribution $Exp(\lambda)$, find the method of moments estimate of $\lambda$

```{r}
t <- c(5, 11, 21, 31, 46, 75, 98, 122, 145, 165, 196, 224, 245, 293, 321, 330, 350, 420)
```


## A uniform mystery

Suppose $X_1, \ldots, X_n \sim Unif(a, b)$, but both $a$ and $b$ are unknown. 

a. Construct simultaneous method of moment estimators $\hat{a}$ and $\hat{b}$ from the facts that $E(X)=\frac{a+b}{2}$ and $Var(X)=\frac{(b-a)^2}{12}$.

b. One could also use the estimator $\hat{b}=\frac{n+1}{n}\max(X_1, \ldots, X_n)$, which is unbiased. Show that $\frac{n+1}{n}(\max(X)-\min(X))$ is an unbiased estimator for $b-a$ using Monte Carlo.

c. From Part b, use algebra to figure out an estimator $\hat{a}$ for the lower bound of the support.

d. Is the estimator from part (c) unbiased? Is it possible that $\hat{a} > \min(X)$?

## Another Uniform Estimation

Suppose $X_1, \ldots X_n \sim Unif(-\theta, theta)$ are an independent random sample. 

a. Use the method of moments to construct an estimator for $\theta$ (hint: the first moment will be useless!)

b. Construct another estimator based on the sample max and sample min. Is it biased?

## Goldfish

In 1997 Pepperidge Farm started putting faces on some of the goldfish crackers they produce. Roughly 40% of goldfish crackers have faces.

Suppose you want to estimate this proportion, and you buy a 20 pack box as a representative sample of Goldfish. The box contains 20 1-oz packs of goldfish, each containing roughly 50 goldfish. Suppose we can model the number of goldfish in a pack using a normal distribution (rounded to the nearest integer), $Y_i \sim N(50, 2^2)$.

To estimate the proportion of smiley faces, you figure you could do three things:

1. Open all 20 bags in one pile and calculate the sample proportion
2. Calculate the proportion in each of the 20 bags and average those proportions
3. Calculate the proportions in each bag and calculate the **weighted** average (weighting by the number of goldfish in each bag)

Using a normal model for the number of fish in a bag, and a binomial for the number of smiley fish, determine the bias and variance of these three estimators.

```{r}
NMC <- 10000
estimates <- data.frame(method1=numeric(NMC),
                        method2=numeric(NMC),
                        method3=numeric(NMC))
for(i in 1:NMC){
  Y <- round(rnorm(20, 50, 2))
  X <- rbinom(20, Y, .40)
  estimates[i,] <- c(sum(X)/sum(Y),
                     mean(X/Y),
                     sum(X/Y * Y/sum(Y))) 
}

colMeans(estimates)
apply(estimates, 2, FUN=var)
```
When working out the math, it becomes clear that method 1 and 3 are actually identical to each other. 
Running the MC a few times it seems that all 3 methods give an unbiased estimate. Method 2 however has a slightly larger variance. This seems consistent after re-running the MC multiple times. Thus method 1 is probably the best method to estimate $p$.

# Beyond STAT 340

These problems are excellent practice but they are beyond the material we cover in STAT 340.


