---
title: "R11: Ridge and Lasso"
author: "Brian Powers"
date: "2023-11-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Required Functions

```{r, message=F}
source("r_scripts/MS_CV_functions.R")
```

## Ridge Regression Example

```{r}
library(MASS);

lambda_vals <- seq(0,10,1)^2; # Choose lambdas to try.
# lm.ridge needs:
# 1) a model (mpg~. says to model mpg as an intercept
#         plus a coefficient for every other variable in the data frame)
# 2) a data set (mtcars, of course)
# 3) a value for lambda. lambda=0 is the default,
#         and recovers classic linear regression.
#         But we can also pass a whole vector of lambdas, like we are about to do,
#         and lm.ridge will fit a separate model for each.
# See ?lm.ridge for details.
ridge_models <- lm.ridge(mpg~., mtcars, lambda=lambda_vals);

# Naively plotting this object shows us how the different coefficients
# change as lambda changes.
plot( ridge_models);

kfoldCV.ridge(32, lambda_vals, mtcars, "mpg")

## Examine the coefficients as lambda increases
as.data.frame(ridge_models$coef)
```

## Ridge Regression With Standardized Data

One problem that we will encounter with Ridge regression (and LASSO) is that it assumes large betas are large because the predictor is important. This is not necessarily true - it could be due to the units of the variable.

Consider a model of mpg based on weight in pounds.
```{r}
lm(mpg~ I(1000*wt), data=mtcars)
```
When we convert weight to pounds (multiply by 1000) then the coefficient is -.005. But if we have wt in 1000s of pounds
```{r}
lm(mpg~ wt, data=mtcars)
```
The coefficient is 1000x as large! For this reason it is a good idea to standardize your data before you apply a shrinkage method. Standardized data requires subtracting the mean and dividing by standard deviation (of that column):
$$ Z_{i,j} = \frac{X_{i,j} - \bar{X}_j}{S_{j}}$$

`lm.ridge` standardizes predictors by default. That's a good thing. WE don't have to worry about standardizing.

## LASSO Regression

#LASSO mpg on mtcars

the function `glmnet` also standardizes predictors by default, so you don't have to worry about it. 

```{r}
lambda_vals <- c(0.001,.1,.2,.3,.5,.7, 1, 1.5, 2, 2.5, 3, 3.5, 4, 5)
kfoldCV.LASSO(32, lambda_vals, mtcars.std, "mpg")
LASSO.fits <- glmnet(mtcars[,-1], mtcars[,1], alpha=1, lambda=lambda_vals)
plot(LASSO.fits, label=TRUE, xvar="lambda", xlim = rev(range(-log(LASSO.fits$lambda))))

LASSO_coeff<- t(coef(LASSO.fits))
as.data.frame(as.matrix(LASSO_coeff))
```


## Boston Dataset Example

```{r, warning=F}
data("Boston")

lambda_vals <- seq(0,5,length.out=10)^2; # Choose lambdas to try.
ridge_models <- lm.ridge(medv ~ . , Boston, lambda=lambda_vals);
plot( ridge_models );
kfoldCV.ridge(4, lambda_vals, Boston, "medv")
```

```{r}
#LASSO 
lambda_vals <- seq(0.0001,.05,length.out=10)

LASSO.fits <- glmnet(Boston[,names(Boston)!="medv"], Boston[,"medv"], alpha=1, lambda=lambda_vals)
plot(LASSO.fits, label=TRUE, xvar="lambda")

kfoldCV.LASSO(10, lambda_vals, Boston, "medv")

resultsTable <- t(as.matrix(coef(LASSO.fits)))
row.names(resultsTable) <- sort(lambda_vals,decreasing=TRUE)
as.data.frame(round(resultsTable,3))
```



